An Analysis of the Feasibility and Performance of BitTorrent as a Data Transfer Protocol in Peer to Peer Backup Systems

### 2014-04-25 paper/paper.tex
Paper: small changes

% TODO make title page look nicer
\begin{titlepage}
\begin{center}
\thispagestyle{empty}
% TODO install \textsc font
\Large{\textbf{A Major Qualifying Project Report\\ \large{ON}}}\\[0.7cm]
\LARGE{\textsc {\textbf{An Analysis of the Feasibility and Performance of BitTorrent as a Data Transfer Protocol in Peer to Peer Backup Systems}}}\\[0.5cm]
\vspace{0.2cm}
\Large{\textbf{\\Submitted to the Faculty of }}
\LARGE{\textbf{\\WORCESTER POLYTECHNIC INSTITUTE\\}}
\vspace{0.5cm}
\Large{\textbf{\\In Partial Fulfillment of the Requirement for the}}
\Large{\textbf{\\Degree of Bachelor of Science}}
\vspace{0.5cm}
\Large{\textbf{\\by}}\\[0.5cm]
\large{\textbf{Dan Bouffard}}\\
\large{\textbf{Marc Green}}\\
\vspace{0.5cm}
\large{\textbf{UNDER THE GUIDANCE OF}}\\
\large{\textbf{Professor G\'abor N. S\'ark\"ozy}}\\
\vspace{1cm}
\large{\textbf{\\May 1, 2014}}\\
\end{center}
\end{titlepage}

@@

Most backup systems that are in use today are centralized, meaning that they rely on some central entity to store their data. This entity is typically a company that offers data backup as a service, such as Dropbox or Google. While these solutions are somewhat convenient, since the service is guaranteed to almost always be online, there are several problems with this approach. The amount of data that users can backup is severely limited, and obtaining more space costs money. In a time where owning terabytes of data is common, cloud storage meant for everyday use is not a feasible backup solution. In addition to this, users must trust the entity to which they send their data to both remain online and respect their privacy. This trust is not always possible, especially in the post-Snowden era, where users must also trust the government under which the central entity resides.
% TODO citation for central backup entity going offline due to govt (bitcloud or something?)
% TODO citation for central backup entity violating trust (google + nsa)

@@

We have designed and implemented a new P2P backup system, called BTBackup, that improves the speed of backing up and recovering data. Our main contribution, and the technology responsible for faster file backup and recovery, is the way in which we handle data transfer: BitTorrent Sync, a new synchronization application created by BitTorrent, Inc. BitTorrent Sync uses a modified version of the BitTorrent protocol in order to quickly synchronize files between groups of peers. This protocol is especially effective at transferring large files, which makes its use perfect for the domain of P2P file backup. By leveraging BitTorrent Sync, we take advantage of BitTorrent's ability to better utilize a peer's bandwidth by downloading from multiple sources at once. This is in contrast to other P2P backup systems, which directly download from one source at a time. Our approach takes advantage of the fact that P2P backup systems require data replication, the act of inserting several redundant copies of data into the network, to combat node churn. We view the the set of nodes storing replications for a given file as a BitTorrent swarm, allowing us to apply the BitTorrent protocol without additional overhead.

@@

\begin{quote}
\begin{itemize}
\item Local peer discovery. All peers inside local network are discovered by sending broadcast packets. If there are peers with the same secret they respond to the broadcast message and connect.
\item Peer exchange (PEX). When two peers are connected, they exchange information about other peers they know.
\item Known hosts (folder settings). If you have a known host with a static ip:port, you can specify this in Sync client, so that it connects to the peer using this information.
\item DHT. Sync uses DHT to distribute information about itself and obtain the information about other peers with this secret. Sync sends SHA1(Secret):ip:port to DHT to announce itself and will get a list of peers by asking DHT for the following key SHA1(Secret)
\item BitTorrent tracker. BitTorrent Sync can use a specific tracker server to facilitate peer discovery. The tracker server sees the combination of SHA1(secret):ip:port and helps peers connect directly. The BitTorrent Sync tracker also acts like a STUN server and can help do a NAT traversal for peers so that they can establish a direct connection even behind a NAT.
\end{itemize}
\end{quote}

In our system, we take advantage of the "known hosts" discovery method. When one peer picks another as a site for a replica of one of its files, it adds that other peer's IP address and listening port to the list of known hosts. When the backups need to be restored, the list can be reloaded into BitTorrent Sync. We decided to use this because it does not rely on the use of a central authority.

@@

@@

There are two instances within this test scenario. The first is our control, where all peers joined the network at the start and did not leave until the end of the instance. Running this test allowed us to see how much data is transferred without any churn. In the second instance, nodes entered and left the network at a rate specified by a distribution function (described in the next paragraph). By having nodes limit their availability, the peers that store data on them perceive them as unreliable and move their replicas to other nodes, creating more network traffic in the process. By measuring this network traffic and comparing it to the traffic of the test instance with no churn, we can determine the average amount of network traffic generated by data layer maintenance.

% TODO: Make sure this is very understandable.
In our test instance with churn, we modeled a network where each peer had an average uptime of 150 seconds (2.5 minutes) and an average downtime of 75 seconds (1.25 minutes). We chose these numbers based on the amount of time a typical user might be online in a given day. We reasoned that, if a typical user keeps his or her computer on during the day and shuts it off when he or she sleeps (for 8 hours), then the uptime in the average day will be 16 hours. As explained in section \ref{subsec:ChallengeMechanism_sec:Fairness_chap:BTBackup}, our challenging mechanism sends a challenge message to backup nodes every 6 hours and blacklists them upon 50\% challenge failure over 4 days, meaning that a node is blacklisted if they fail 8 or more tests. We scaled these numbers down so that we would have a 1 hour test (a factor of 360): a 6 hour challenge becomes a 60 second challenge, and the challenge period becomes 16 minutes. To make the challenge period evenly fit into the 1 hour test, we rounded the challenge period down to 15 minutes, making the scaling a factor of 384. From this factor, we scale 16 hours and 8 hours down to 2.5 minutes and 1.25 minutes, respectively.

@@

To make this test scenario more realistic, each node has their upload and download bandwidth limited to different values at the beginning of each instance. Download speeds are picked from a normal distribution of Internet speeds. In order to have a mean and standard deviation for our distribution that is reflective of current Internet speeds around the world, we found statistics on the average download speed per country, as well as the number of people in each country who use the Internet. This allowed us to find the expected value for download speed by taking a weighted average, $W$:
$$
W=\sum\limits_{i=1}^{|\mathcal{D}|} n_iD_i
$$
Where $\mathcal{D}$ is the set of country download speeds and $n$ is the corresponding percentages of global Internet users that each country has. The normal distribution is only used for download speeds; rather than choosing an upload speed using the same method, we took each download speed and halved it. This ensures that the upload speed is lower than the corresponding download speed, which is often the case on the Internet, as well as ensuring that the two speeds are proportional.

Our distributions are a combination of three functions over the range $[0,1)$. In order to choose an uptime or downtime, a node chooses a random number in this range, and applies it to the following piecewise function:

$$
f(x)=
\begin{cases}
10xc & \mbox{ if $0 \leq x < 0.2$ } \\
[4(x-0.2)+2]c & \mbox{ if $0.2 \leq x < 0.8$ } \\
[e^{21(x-0.8)}+3.4]c & \mbox{ if $0.8 \leq x < 1$ }
\end{cases}
$$

The value of $c$ is what differentiates between choosing an uptime or a downtime. In order to obtain the average uptime and downtime stated above, we used $c=50$ for uptime and $c=25$ for downtime. Since the middle equation is our biggest range, we modeled it so that its average values for each value of $c$ approximately corresponds to the average uptime and downtime. We used a steeper linear equation from 0 to 0.2 so that very short uptimes and downtimes are represented, but very few nodes will ever choose values using this function. The steepness also makes the transition to the middle equation quick but continuous. For the top range, we chose an exponential function so very long uptimes and downtimes are represented. For example, the upper bound for uptime is 3504.32 seconds, or 58.41 minutes for $c=50$, which is almost the entire length of the test. Nodes who pick these values represent computers that are almost always online, such as servers or NAS units.

@@

@@

We designed this scenario to simulate an average level of churn over 16 days. The difference in data sent between the instance with churn and the one without is 14.27 GB. This means, on average, each peer sent an additional 142.7 MB of data in the presence of churn. In other words, with our system, the average peer can expect to create an additional replica every $\dfrac{16 \text{ days}}{142.7 \text{ MB}} \times 100 \text{ MB per replica} = 11.21 \text{ days}$.
% TODO not sure what to do with this number. is this a lot? it may be if the user is backing up terabytes of data

>>>>

% TODO make title page look nicer
\begin{titlepage}
\begin{center}
\thispagestyle{empty}
% TODO install \textsc font
\Large{\textbf{A Major Qualifying Project Report\\ \large{ON}}}\\[0.7cm]
\LARGE{\textsc {\textbf{Improving the Speed of Peer to Peer Backup Systems with BitTorrent}}}\\[0.5cm]
\vspace{0.2cm}
\Large{\textbf{\\Submitted to the Faculty of }}
\LARGE{\textbf{\\WORCESTER POLYTECHNIC INSTITUTE\\}}
\vspace{0.5cm}
\Large{\textbf{\\In Partial Fulfillment of the Requirement for the}}
\Large{\textbf{\\Degree of Bachelor of Science}}
\vspace{0.5cm}
\Large{\textbf{\\by}}\\[0.5cm]
\large{\textbf{Dan Bouffard}}\\
\large{\textbf{Marc Green}}\\
\vspace{0.5cm}
\large{\textbf{UNDER THE GUIDANCE OF}}\\
\large{\textbf{Professor G\'abor N. S\'ark\"ozy}}\\
\vspace{1cm}
\large{\textbf{\\May 1, 2014}}\\
\end{center}
\end{titlepage}

@@

Most backup systems that are in use today are centralized, meaning that they rely on some central entity to store their data. This entity is typically a company that offers data backup as a service, such as Dropbox or Google. While these solutions are somewhat convenient, since the service is guaranteed to almost always be online, there are several problems with this approach. The amount of data that users can backup is severely limited, and obtaining more space costs money. In a time where owning terabytes of data is common, cloud storage meant for everyday use is not a feasible backup solution. Dedicated backup solutions can offer an adequate amount of space, but often require expensive recurring payments for their service. In addition to this, users must trust the entity to which they send their data to both remain online and respect their privacy. This trust is not always possible, especially in the post-Snowden era, where users must also trust the government under which the central entity resides.
% TODO citation for central backup entity going offline due to govt (bitcloud or something?)
% TODO citation for central backup entity violating trust (google + nsa)

@@

We have designed and implemented a new P2P backup system, called \mbox{BTBackup}, that improves the speed of backing up and recovering data. Our main contribution, and the technology responsible for faster file backup and recovery, is the way in which we handle data transfer: BitTorrent Sync, a new synchronization application created by BitTorrent, Inc. BitTorrent Sync uses a modified version of the BitTorrent protocol in order to quickly synchronize files between groups of peers. This protocol is especially effective at transferring large files, which makes its use perfect for the domain of P2P file backup. By leveraging BitTorrent Sync, we take advantage of BitTorrent's ability to better utilize a peer's bandwidth by downloading from multiple sources at once. This is in contrast to other P2P backup systems, which directly download from one source at a time. Our approach takes advantage of the fact that P2P backup systems require data replication, the act of inserting several redundant copies of data into the network, to combat node churn. We view the the set of nodes storing replications for a given file as a BitTorrent swarm, allowing us to apply the BitTorrent protocol without additional overhead.

@@

% TODO why do we discuss the different ways to find peers? We only need to talk about predefined hosts
\begin{quote}
\begin{itemize}
\item Local peer discovery. All peers inside local network are discovered by sending broadcast packets. If there are peers with the same secret they respond to the broadcast message and connect.
\item Peer exchange (PEX). When two peers are connected, they exchange information about other peers they know.
\item Known hosts (folder settings). If you have a known host with a static ip:port, you can specify this in Sync client, so that it connects to the peer using this information.
\item DHT. Sync uses DHT to distribute information about itself and obtain the information about other peers with this secret. Sync sends SHA1(Secret):ip:port to DHT to announce itself and will get a list of peers by asking DHT for the following key SHA1(Secret)
\item BitTorrent tracker. BitTorrent Sync can use a specific tracker server to facilitate peer discovery. The tracker server sees the combination of SHA1(secret):ip:port and helps peers connect directly. The BitTorrent Sync tracker also acts like a STUN server and can help do a NAT traversal for peers so that they can establish a direct connection even behind a NAT.
\end{itemize}
\end{quote}

@@

% TODO talk about specific numbers used to deem unreliability

@@

There are two instances within this test scenario. The first is our control, where all peers joined the network at the start and did not leave until the end of the instance. Running this test allowed us to see how much data is transferred without any churn. In the second instance, nodes enter and leave the network at a rate specified by the distribution function described below. % Pickup here

% TODO: Make sure this is very understandable.
In our tests, we modeled a network where each peer had an average uptime of 150 seconds (2.5 minutes) and an average downtime of 75 seconds (1.25 minutes). We chose these numbers based on the amount of time a typical user might be online in a given day. We reasoned that, if a typical user keeps his or her computer on during the day and shuts it off when he or she sleeps (for 8 hours), then the uptime in the average day will be 16 hours. As explained in section \ref{subsec:ChallengeMechanism_sec:Fairness_chap:BTBackup}, our challenging mechanism sends a challenge message to backup nodes every 6 hours and blacklists them upon 50\% challenge failure over 4 days, meaning that a node is blacklisted if they fail 8 or more tests. We scaled these numbers down so that we would have a 1 hour test (a factor of 360): a 6 hour challenge becomes a 60 second challenge, and the challenge period becomes 16 minutes. To make the challenge period evenly fit into the 1 hour test, we rounded the challenge period down to 15 minutes, making the scaling a factor of 384. From this factor, we scale 16 hours and 8 hours down to 2.5 minutes and 1.25 minutes, respectively.

@@

Our distributions are a combination of three functions over the range $[0,1)$. In order to choose an uptime or downtime, a node chooses a random number in this range, and applies it to the following multipart function:

$$
f(x)=
\begin{cases}
10xc & \mbox{ if $0 \leq x < 0.2$ } \\
[4(x-0.2)+2]c & \mbox{ if $0.2 \leq x < 0.8$ } \\
[e^{21(x-0.84)}+3.5]c & \mbox{ if $0.8 \leq x < 1$ }
\end{cases}
$$

% TODO: Is it accurate to say that the multipart function is continuous?
The value of $c$ is what differentiates between choosing an uptime or a downtime. In order to obtain the average uptime and downtime stated above, we used $c=50$ for uptime and $c=25$ for downtime. Since the middle equation is our biggest range, we modeled it so that its average values for each value of $c$ approximately corresponds to the average uptime and downtime. We used a steeper linear equation from 0 to 0.2 so that very short uptimes and downtimes are represented, but very few nodes will ever choose values using this function. The steepness also makes the transition to the middle equation quick but continuous. For the top range, we chose an exponential function so very long uptimes and downtimes are represented. For example, the upper bound for uptime is 3509.32 seconds, or 58.49 minutes for $c=50$, which is almost the entire length of the test. Nodes who pick these values represent computers that are almost always online, such as servers or NAS units.

@@

% TODO: Talk about what we actually do, including the control group of no churn.

% TODO discuss what we want to get out of this scenario, like the first paragraph in other scenario

% Commenting this out b/c we don't do this test
%\subsection{Number of Replicas} \label{subsec:NumberofReplicas_sec:TestScenarios_chap:Methodology}

% TODO: Talk about the exact distributions used for the constant churn in this scenario.
%This test scenario produced data so that we could determine how the number of replicas influences the recoverability of backed up files, as well as the speed of backup recovery. This is done in the presence of some constant amount of churn, in order to make the tests more realistic. Our testing procedure was very similar to the previous scenario: we created a network with 100 peers that backed up 1GB each. Instead of changing the churn rate between test instances, however, we changed the number of replicas that each peer attempts to find when creating a backup of its file. We started with 3 replicas per file in the first test instance and increased that number to 10 in the last instance. By having this much variation, we were able to see how much of a gain or loss in performance existed when replicas were added or removed. From this, we were able to see which replica counts do not offer enough of a performance gain to warrant the creation of that many replicas. This is especially important to our system, because as a peer creates more replicas, it is required to store more data from other peers.

@@

We designed this scenario to simulate an average level of churn over 16 days. The difference in data sent between the instance with churn and the one without is 14.27 GB. This means, on average, each peer sent an additional 142.7 MB of data in the presence of churn. In other words, with our system, the average peer can expect to replace a replica every $\dfrac{16 \text{ days}}{142.7 \text{ MB}} \times 100 \text{ MB per replica} = 11.21 \text{ days}$.
% TODO not sure what to do with this number. is this a lot? it may be if the user is backing up terabytes of data
